Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2015-03-13 manuscript/ms.tex
Added figures to manuscript

@@

\usepackage{amsmath}
\usepackage{amssymb}
\pdfoutput=1
\usepackage[pdftex]{graphicx}
%\usepackage[mathbf,mathcal]{euler}
%\usepackage{citesort}
\usepackage{dcolumn}
\usepackage{boxedminipage}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[group-separator={,}]{siunitx}

% The figures are in a figures/ subdirectory.
% \graphicspath{{figures/}}

@@

\begin{figure*}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{pipeline2}
  \caption{{\bf Diagrammatic representation of the various stages of the Ensembler pipeline.}
  The number of viable models surviving each stage of the pipeline for are shown, either for all tyrosine kinases (\emph{All TKs}) or representative individual kinases (\emph{SRC} and \emph{ABL}).
  In addition, the typical timing on a cluster (containing Intel Xeon E5-2665 2.4GHz hyperthreaded processors and NVIDIA GTX-680 or GTX-Titan GPUs) is reported to convey resources required per model and for modeling the entire set of tyrosine kinases.
  Note that \emph{CPU-h} denotes the number of hours consumed by the equivalent of a single hyperthread---parallel execution can reduce wall clock time nearly linearly.
  }
  \label{figpipeline}
\end{figure*}

@@

While protein-only models may be sufficient for structural analysis or implicit solvent simulations, {\bf Ensembler} also provides a stage for solvating models with explicit water and performing a round of explicit-solvent MD refinement/equilibration under isothermal-isobaric (NPT) conditions.
The solvation step solvates each model for a given target with the same number of waters to facilitate the integration of data from multiple simulations, such as the construction of MSMs.
The target number of waters is selected by first solvating each model with a specified padding distance (default: 10 \AA), then taking a percentile value from the distribution (default: 68th percentile).
{\color{red}[JDC: Would be useful to explain why we are doing this.]}
{\color{blue}[DLP: Addressed.]}
This helps to prevent models with particularly long, extended loops---such as those arising from template structures with unresolved termini---from imposing very large box sizes on the entire set of models.
Models are resolvated with the target number of waters by first solvating with zero padding, then incrementally increasing the box size and resolvating until the target is exceeded, then finally deleting sufficient waters to match the target value.
The explicit solvent MD simulation is also implemented using OpenMM, using the Amber99SB-ILDN force field and TIP3P water {\color{red}[JDC: CITE]} by default.
Other force fields or water models such as TIP4P-Ew {\color{red}[CITE]}) can be specified via the {\bf Ensembler} API.
{\color{red}[JDC: We should allow other water models in OpenMM too, such as TIP4P-Ew?]}
{\color{blue}[DLP: I forgot to mention this in the text previously - any of the OpenMM force fields can be chosen via the API. I've updated the text accordingly. Is this functionality sufficient? I guess it's ok to leave ff choice as an "advanced" feature which requires use of the API? Otherwise I could add a --water\_model flag to the CLI, for example.]}

@@

% \begin{table}
% \begin{tabular}{l|c|c|c}
%
%     & SRC & ABL & All 90 TKs \\
%     \hline
%     Templates & \num{4433} & \num{4433} & \num{398970} \\
%     Models & \num{4248} & \num{4250} & \num{382568} \\
%     Unique models & \num{4093} & \num{4247} & \num{378839} \\
%     Refined models & \num{4023} & \num{4241} & \num{373513} \\
%
% \end{tabular}
% \label{tab:model_counts}
% \caption{Model counts for the kinase domains of human SRC and ABL, and for all of the 90 human TK targets, at various stages of the Ensembler pipeline.}
% \end{table}
%
% \begin{table}
% \begin{tabular}{l|c}
%
%     & Timing ($h$:$m$:$s$) \\
%     \hline
%     Template reconstruction & 03:00:00 \\
%     Modeling & 00:01:19 \\
%     Refinement (100 ps implicit solvent MD) & 00:01:58 \\
%
% \end{tabular}
% \label{tab:timings}
% \caption{Average timings per template/model for the compute-intensive stages of the Ensembler pipeline.}
% \end{table}

>>>>

\pdfoutput=1
\usepackage[pdftex]{graphicx}

@@

\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[mathbf,mathcal]{euler}
%\usepackage{citesort}
\usepackage{subcaption}
\usepackage{dcolumn}
\usepackage{boxedminipage}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[group-separator={,}]{siunitx}

% The figures are in a figures/ subdirectory.
\graphicspath{{../figures/}}

@@

\begin{figure*}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{pipeline/pipeline2}
  \caption{{\bf Diagrammatic representation of the various stages of the Ensembler pipeline.}
  The number of viable models surviving each stage of the pipeline for are shown, either for all tyrosine kinases (\emph{All TKs}) or representative individual kinases (\emph{SRC} and \emph{ABL}).
  In addition, the typical timing on a cluster (containing Intel Xeon E5-2665 2.4GHz hyperthreaded processors and NVIDIA GTX-680 or GTX-Titan GPUs) is reported to convey resources required per model and for modeling the entire set of tyrosine kinases.
  Note that \emph{CPU-h} denotes the number of hours consumed by the equivalent of a single hyperthread---parallel execution can reduce wall clock time nearly linearly.
  }
  \label{figpipeline}
\end{figure*}

@@

While protein-only models may be sufficient for structural analysis or implicit solvent simulations, {\bf Ensembler} also provides a stage for solvating models with explicit water and performing a round of explicit-solvent MD refinement/equilibration under isothermal-isobaric (NPT) conditions.
The solvation step solvates each model for a given target with the same number of waters to facilitate the integration of data from multiple simulations, such as the construction of MSMs.
The target number of waters is selected by first solvating each model with a specified padding distance (default: 10 \AA), then taking a percentile value from the distribution (default: 68th percentile).
{\color{red}[JDC: Would be useful to explain why we are doing this.]}
{\color{blue}[DLP: Addressed.]}
This helps to prevent models with particularly long, extended loops---such as those arising from template structures with unresolved termini---from imposing very large box sizes on the entire set of models.
Models are resolvated with the target number of waters by first solvating with zero padding, then incrementally increasing the box size and resolvating until the target is exceeded, then finally deleting sufficient waters to match the target value.
The explicit solvent MD simulation is also implemented using OpenMM, using the Amber99SB-ILDN force field and TIP3P water {\color{red}[JDC: CITE]} by default.
Other force fields or water models such as TIP4P-Ew {\color{red}[CITE]}) can be specified via the {\bf Ensembler} API.
{\color{red}[JDC: We should allow other water models in OpenMM too, such as TIP4P-Ew?]}
{\color{blue}[DLP: I forgot to mention this in the text previously - any of the OpenMM force fields can be chosen via the API. I've updated the text accordingly. Is this functionality sufficient? I guess it's ok to leave ff choice as an "advanced" feature which requires use of the API? Otherwise I could add a -{}-water\_model flag to the CLI, for example.]}

@@

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{seqid_dist/seqid_dist}
    \caption{{\bf Sequence identity distribution for human TK models.}
    Distribution of sequence identities for all 373,513 models generated for the human tyrosine kinases.
    {\color{blue}[DLP: should I mention the use of KDE smoothing?]}
    Sequence identities are calculated from pairwise target-template alignments.
    The cumulative distribution function is shown by the dashed line.
    }
  \label{figseqid_dist}
\end{figure}

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{rmsddist/rmsddist2}
    \caption{{\bf RMSD distribution by sequence identity.}
    RMSD distributions for all 373,513 human TK models, divided into three sequence identity ranges.
    % TODO may want to mention KDE smoothing
    For a given target, model RMSDs are calculated relative to the highest sequence identity model for that target.
  }
  \label{figrmsddist}
\end{figure}

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{superposition-src_abl/superposed-seqid_classes-clustered-one_fig}
    \caption{{\bf Superposition of clustered models of Src and Abl.}
    Superposed renderings of nine models each for Src and Abl, giving some indication the diversity of conformations generated by Ensembler.
    The models for each target were divided into three sequence identity ranges (as in Fig. \ref{figrmsddist}), and RMSD-based $k$-medoids clustering was performed to select three clusters from each.
    The models shown are the centroids of each cluster.
    Models are colored and given transparency based on their sequence identity, so that high sequence identity models are blue and opaque, while lower sequence identity models are transparent and red.
  }
  \label{figsuperposition}
\end{figure}

\begin{figure}
  \centering
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=1.0\textwidth]{residue_pair_distances/src/distances}
      \caption{Src}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
      \includegraphics[width=1.0\textwidth]{residue_pair_distances/abl/distances}
      \caption{Abl}
    \end{subfigure}
    \caption{{\bf Residue pair distances for Src and Abl.}
    % ...
  }
  \label{figaa_pair_distances_ref_structures}
\end{figure}

\begin{figure}[tb]
  \centering
    \includegraphics[width=1.0\textwidth]{residue_pair_distances/src/src_ref_structures}
    \caption{{\bf Reference structures for Src.}
    % ...
  }
  \label{figsrc_ref_structures}
\end{figure}
